---
title: "ADSC2030 Project Report"
author: "Neha Malhan (T00757554) & Thiwanka Abeyweera (T00754871)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: true
    number_sections: true
  word_document: default
header-includes: \usepackage{setspace}
---

\newpage 
\tableofcontents 
\listoffigures
\listoftables

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries,echo=FALSE,warning=FALSE,message=FALSE}

library(knitr)
library(dplyr)
library(lsmeans)
library(car)
library(lme4)

```

# Introduction (1 page)

## Discuss the data (Briefly discuss data)
 

## Data Description

*Datasets for this study were obtained from an official website of the United States government, National Institute of Standards and Technology. The website contains a database of results from many different experiments following different experimental designs from which a study conducted on studying survival time of poison antidote and the acidity level of sour cream were selected.*

The Poison-Antidote dataset comes from an experimental study designed to assess the effect of various antidotes on animal survival following exposure to different poisons. The study follows a Complete Randomized Block Design (CRBD) with one factor, where Poison serves as the blocking factor and is represented by three blocks, each containing a set of animals exposed to the same poison type or concentration. Antidote is the treatment factor, which has four levels corresponding to different antidotes tested for their effectiveness in counteracting the toxic effects of the poison.

The Survival Time of animals is the response variable, and the study aims to determine how different antidotes influence survival after poison exposure. Each antidote is randomly assigned to animals within each block, and each antidote is replicated three times per block. The dataset includes 48 observations, providing sufficient data for statistical analysis and a comprehensive evaluation of the interaction between poison and antidote treatments on animal survival.

### **Study Variables:**

-   **Survival Time**: The time (in hours) that an animal survives after exposure to a specific poison and subsequent antidote treatment.

-   **Poison**: The type or concentration of poison administered to the animals (3 levels).

-   **Antidote**: The treatment administered to the animals after poison exposure (4 levels).

```{r data,echo=FALSE, results='hide'}

Poison_Antidote <- read.csv("Poison Antidote_RCB.csv")

colnames(Poison_Antidote) <- c("Poison", "Antidote", "Rep", "SurvivalTime")

Poison_Antidote$Poison <- as.factor(Poison_Antidote$Poison)
Poison_Antidote$Antidote <- as.factor(Poison_Antidote$Antidote)
Poison_Antidote$Rep <- as.factor(Poison_Antidote$Rep)

Sour_Cream <- read.csv("Acidity of Sour Cream_LSD.csv")

Sour_Cream$Batch <- as.factor(Sour_Cream$Batch)
Sour_Cream$Position <- as.factor(Sour_Cream$Position)
Sour_Cream$Starter <- as.factor(Sour_Cream$Starter)
```

```{r displaypoi,echo=FALSE}

kable(head(Poison_Antidote), caption = "Poison-Antidote Datset")
```

The Acidity of Sour Cream dataset originates from an experimental study examining the factors influencing the acidity levels in sour cream production. The experiment has followed a 5-level Latin Square Design, incorporating three treatment factors: batch, position, and starter culture. The dataset contains 25 observations.

From each of the five batches in the study, samples have been collected from five different positions. Each sample was then assigned one of five starter cultures, coded numerically from 1 to 5. Since the dataset follows a 5-level Latin Square Design, each starter culture appears once per row and once per column, controlling for batch and position effects. As there are no direct replicates within the dataset, the statistical power of the analysis relies on the structure of the experimental design rather than traditional replication.

### Study Variables:

-   **Acidity:** The response variable, representing the measured acidity levels of sour cream samples.

-   **Batch:** A categorical factor representing different production batches (5 levels).

-   **Position:** The location within the batch where the sample was taken (5 levels).

-   **Starter Culture:** The type of culture used in fermentation (5 levels: C = 1, H = 2, M = 3, N = 4, Y = 5).

```{r displaysour,echo=FALSE}

kable(head(Sour_Cream), caption = "Acidity of Sour Cream Datset")
```

## Objective (State the overall objective(s) of the experiment(s).)

This study comprise of two main sections. The objective of the first section is to examine the effects of different antidotes and poisons on survival time. The study aims to determine the impact of antidote and poison types on survival outcomes while assessing potential interaction effects between these factors. In addition, which factor (antidote or poison) has the most significant influence on survival time will be evaluated.

*The second section of the study focuses on variance analysis using the Acidity of Sour Cream dataset. This analysis will explore whether acidity levels vary significantly across different batches, the position within a batch, or the type of starter culture used. In addition, we will assess whether acidity variance remains consistent across all groups.*

# Research Objectives (2 pages)

-   Clearly state the research objective(s) (officially)

-   State the hypothesis/hypotheses.

-   May include multiple research questions/hypotheses (state them all here).

## Research Question 1

**Hypotheses**:

*Null Hypothesis (*$H_0$) : Mean body fat percentage is the same for both age groups ($\mu_{\text{< 40}} = \mu_{\geq 40}$)

*Alternate Hypothesis (*$H_1$) : Mean body fat percentage differs between the two age groups ($\mu_{\text{< 40}} \neq \mu_{\geq 40}$)

## Research Question 2

**Hypotheses:**

*Null Hypothesis (*$H_0$) : The mean body fat percentages are equal across all BMI categories

($\mu_{\text{underweight}} = \mu_{\text{normal weight}} = \mu_{\text{overweight}} = \mu_{\text{obesity}}$)

*Alternate Hypothesis (*$H_1$) : At least one BMI category has a different mean body fat percentage

# Analysis Selection (2-3 pages) 

Which tools will you use to analyze data and why?

# Results (4-5 pages) 

Clearly present results including effects estimates, hypothesis tests about the model(s), ANOVA tables, diagnostics, etc.

Interpret the results (how do these results relate to the experiment(s)’ objectives?)

# Conclusions

# References

1\. National Institute of Standards and Technology. (n.d.). *Datasets*. U.S. Department of Commerce. <https://www.itl.nist.gov/div898/software/dataplot/datasets.htm>

# Appendices

Visualizations such as pair plots, and additional regression diagnostics, along with the code used to generate them, offer additional evidence to support the analysis and results discussed in this report. These materials are included in the appendix for reference.

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

```{r plots}

```

```{r assumption_check, warning=FALSE, message=FALSE}

```
