---
title: "ADSC2030_Project"
author: "Thiwanka & Neha"
format: pdf
editor: visual
---

## Objective 1 - Poison Antidote - CRB

The first section of the study examines how different antidotes and poisons influence survival time. It aims to identify which factor has the greatest impact on survival while also assessing potential interactions between antidote and poison types.

**Data Source and Study Description:**

The Poison-Antidote dataset comes from an experimental study designed to assess the effect of various antidotes on animal survival following exposure to different poisons. The study follows a Complete Randomized Block Design (CRBD) with one factor, where Poison serves as the blocking factor and is represented by three blocks, each containing a set of animals exposed to the same poison type or concentration. Antidote is the treatment factor, which has four levels corresponding to different antidotes tested for their effectiveness in counteracting the toxic effects of the poison.

The Survival Time of animals is the response variable, and the study aims to determine how different antidotes influence survival after poison exposure. Each antidote is randomly assigned to animals within each block, and each antidote is replicated three times per block. The dataset includes 48 observations, providing sufficient data for statistical analysis and a comprehensive evaluation of the interaction between poison and antidote treatments on animal survival.

### **Study Variables:**

-   **Survival Time**: The time (in hours) that an animal survives after exposure to a specific poison and subsequent antidote treatment.

-   **Poison**: The type or concentration of poison administered to the animals (3 levels).

-   **Antidote**: The treatment administered to the animals after poison exposure (4 levels).

```{r}

Poison_Antidote <- read.csv("Poison Antidote_RCB.csv")

colnames(Poison_Antidote) <- c("Poison", "Antidote", "Rep", "SurvivalTime")

Poison_Antidote$Poison <- as.factor(Poison_Antidote$Poison)
Poison_Antidote$Antidote <- as.factor(Poison_Antidote$Antidote)
Poison_Antidote$Rep <- as.factor(Poison_Antidote$Rep)

```

#### 1.1 Determining how antidote and poison affect survival time

The mathematical model for the randomized complete block design (RCB) with one factor that has been applied to the Poison-Antidote dataset is:

$Y_{ij}$ = $\mu$ + $b_i$ + $\tau_j$ + $\epsilon_{ij}$

Where:

-   $Y_{ij}$​ represents the observed survival time for the $j_{th}$ antidote in the $i_{th}$ block,

-   $\mu$ is the overall mean survival time,

-   $b_i$ represents the block effects associated with different poisons,

-   $\tau_j$​ represents the treatment effects of different antidotes,

-   $\epsilon_{ij}$​ is the random error term.

```{r}

# Determining how antidote and poison affect survival time

model1 <- aov(SurvivalTime ~ Poison + Antidote, data = Poison_Antidote)
summary(model1)

```

Both the Poison and Antidote factors exhibit significant effects on survival time. The sum of squares (Sum Sq) reveals that Poison accounts for a greater proportion of the variability in survival time compared to Antidote, as reflected in the higher sum of squares for Poison compared to Antidote).

The F value for Poison suggests that Poison explains a substantial amount of variance in survival time, which is significantly higher than the F value for Antidote. This indicates that the blocking factor (Poison) has a stronger influence on survival time, though Antidote still plays a significant role.

The residuals (error term) show the unexplained variance, which is the highest among the components of the model. Despite the relatively large residual variance, the fact that both Poison and Antidote explain a considerable proportion of the variability in survival time justifies the model's use.

By incorporating Poison as a blocking factor, the model controls for baseline variability associated with individual animals' response to different poisons, which helps isolate the effect of the Antidote treatments. The statistical significance of both factors indicates that they each have a meaningful impact on survival time. Blocking by poison ensures that any residual variability is attributed to the antidote treatment rather than natural differences in animal survival, providing a clearer understanding of how different antidotes influence survival time.

#### 1.2 **The interaction between the blocking factor (Poison) and the treatment factor (Antidote)** and their combined effects on the survival time of animals

```{r}

# Modifying the ANOVA model to include interaction between Antidote and Poison

model2 <- aov(SurvivalTime ~ Poison * Antidote, data = Poison_Antidote)
summary(model2)

```

This model tests the effects of both Poison and Antidote on survival time, as well as their potential interaction. The interaction term between Poison and Antidote is not significant. This means that the effect of the antidote does not depend on the type of poison. In other words, the effectiveness of each antidote is consistent across different poison levels, and there is no significant combined effect between poison and antidote on survival time.

#### 1.3 Examining the nature of the relationship between treatment and response

```{r}

contrasts(Poison_Antidote$Antidote) <- contr.poly(4)

summary.aov(model1, split = list(Antidote = list("Linear" = 1,
              "Quadratic" = 2, "Cubic" = 3)))
```

The analysis shows that the Antidote factor has both linear and cubic effects on survival time, and these effects are statistically significant. The Antidote factor exhibits a significant linear relationship with survival time, meaning that increasing antidote levels consistently affect survival time. Additionally, the cubic relationship is also significant, implying that the effect of antidote is not purely linear and may show more complex behavior at higher levels. However, the quadratic effect is not significant, suggesting that a simple curve is not sufficient to describe the relationship, but higher-order (cubic) terms might better capture the complexity.

The plots below illustrate the relationship between survival time and antidote levels. The data is summarized by calculating the mean survival time for each antidote group, and polynomial regression models (linear and cubic) are fitted to capture the trends in the relationship. The fitted curves are overlaid on the plot for visualization.

```{r}

library(dplyr)  # Use dplyr for proper data handling

# Compute mean SurvivalTime for each Antidote group
df_summary <- Poison_Antidote %>%
  group_by(Antidote) %>%
  summarise(mean_survival = mean(SurvivalTime, na.rm = TRUE))

x <- as.numeric(as.character(df_summary$Antidote))  # Convert Antidote levels to numeric
y <- df_summary$mean_survival  # Correctly computed mean SurvivalTime per Antidote

plot(x, y, xlab = "Antidote", ylab = "SurvivalTime")

xx <- seq(min(x), max(x), length.out = 100)  # Adjust range dynamically
rate.quad <- lm(y ~ poly(x, 3))  # Fit polynomial model

lines(xx, predict(rate.quad, data.frame(x = xx)), col = "blue")  # Draw fitted curve

```

```{r}

# Compute mean SurvivalTime for each Antidote group
df_summary <- Poison_Antidote %>%
  group_by(Antidote) %>%
  summarise(mean_survival = mean(SurvivalTime, na.rm = TRUE))

x <- as.numeric(as.character(df_summary$Antidote))  # Convert Antidote to numeric
y <- df_summary$mean_survival  # Mean SurvivalTime per Antidote

# Check lengths before proceeding
cat("Length of x:", length(x), "\n")
cat("Length of y:", length(y), "\n")

if (length(x) == length(y)) {
  plot(x, y, xlab = "Antidote", ylab = "Survival Time")

  xx <- seq(min(x), max(x), length.out = 100)  # Adjust range dynamically
  rate.quad <- lm(y ~ poly(x, 1))  # Fit polynomial model

  lines(xx, predict(rate.quad, data.frame(x = xx)), col = "blue")  # Draw fitted curve
} else {
  warning("Mismatch in x and y lengths. Model cannot be fitted.")
}

```

#### 1.4 How effective is the blocking in this example?

```{r}

sigma2CRD <- (10330 + 10509)/(4*(3-1))
sigmaRCB <- 250 # From residuals

nuRCB <- 6 #(b-1)(t-1) #b=10 #t=5
nuCRD <- 8 #t(b-1)


RE <- ((nuRCB+1)*(nuCRD+3)*sigma2CRD)/((nuRCB+3)*(nuCRD+1)*sigmaRCB)

cat("The Relative Efficiency (RE) is:", RE, "\n")

```

The Relative Efficiency (RE) compares the Randomized Complete Block Design (RCB) to the Completely Randomized Design (CRD), quantifying the effectiveness of blocking in reducing variance. The estimated variance for the CRD is 2604.875, while the residual variance in the RCB is 250. This indicates that blocking has significantly improved efficiency, reducing residual variance by approximately $1 - \frac{250}{2604.875} \approx 90\%$ compared to a CRD. Without blocking, achieving the same precision in treatment means would require about nine times as many trials. By accounting for poison-related variability, blocking has enhanced the overall efficiency of the experiment.

#### 1.5 Determining the significance of effects

```{r}

## Determining the significance of effects (work for both equal and unequal number of replicates)

library(lsmeans)

lsmeans(model1, pairwise ~ Antidote, adjust = c("tukey"))

```

The $lsmeans()$ results indicate that Antidote 2 has the highest estimated mean survival time, followed by Antidote 4, while Antidote 1 and 3 show relatively lower means. The standard errors and degrees of freedom are consistent across all levels of the antidote factor, ensuring balanced comparisons.

The pairwise contrasts reveal significant differences between Antidote 1 and 2, Antidote 1 and 4, and Antidote 2 and 3, as indicated by their low p-values. The largest difference is observed between Antidote 2 and 3, reinforcing the $lsmeans()$ findings. In addition, the confidence intervals for the $lsmeans()$ support these significant differences, further confirming that Antidote 2 has the strongest effect on survival time.

#### 1.6 Model diagnostics

For the mathematical model used to analyze the RCB with one factor to be valid, the usual assumptions of normality of experimental error and homogeneity of variance across treatment factors and blocks must hold.

```{r}

library(car)

# Normality #
plot(model1,2)

shapiro.test(model1$residuals)

# Constant Variance #
plot(model1,3)

leveneTest(resid(model1) ~ Poison_Antidote$Antidote)
```

The Q-Q plot of residuals suggests that the experimental error deviates from normality, likely due to a few data points that depart from the normal line, which may be potential outliers. This is further confirmed by the Shapiro-Wilk test, which yields a low p-value, indicating a significant departure from normality. The Scale-Location plot does not indicate major violations of the assumption of constant variance. Additionally, the Levene’s test results in a high p-value, suggesting that the variance of experimental error is homogeneous across different treatment factor levels.

Thus, while the assumption of homogeneity of variance holds, the normality assumption is violated, which can be addressed by applying appropriate correction methods.

## Objective 2 - Acidity of Sour Cream - LSD

This analysis investigates how acidity levels vary across batches, sampling positions within a batch, and different starter culture types. It also examines the relative contribution of each factor to overall variability while assessing whether certain sources of variation are more influential than others.

**Data Source and Study Description:**

The Acidity of Sour Cream dataset originates from an experimental study examining the factors influencing the acidity levels in sour cream production. The experiment has followed a 5-level Latin Square Design, incorporating three treatment factors: batch, position, and starter culture. The dataset contains 25 observations.

From each of the five batches in the study, samples have been collected from five different positions. Each sample was then assigned one of five starter cultures, coded numerically from 1 to 5. Since the dataset follows a 5-level Latin Square Design, each starter culture appears once per row and once per column, controlling for batch and position effects. As there are no direct replicates within the dataset, the statistical power of the analysis relies on the structure of the experimental design rather than traditional replication.

### Study Variables:

-   **Acidity:** The response variable, representing the measured acidity levels of sour cream samples.

-   **Batch:** A categorical factor representing different production batches (5 levels).

-   **Position:** The location within the batch where the sample was taken (5 levels).

-   **Starter Culture:** The type of culture used in fermentation (5 levels: C = 1, H = 2, M = 3, N = 4, Y = 5).

```{r}

Sour_Cream <- read.csv("Acidity of Sour Cream_LSD.csv")

Sour_Cream$Batch <- as.factor(Sour_Cream$Batch)
Sour_Cream$Position <- as.factor(Sour_Cream$Position)
Sour_Cream$Starter <- as.factor(Sour_Cream$Starter)

```

### 1.1 Determining how the factors affect the acidity

To study the variance in the response caused by varying the levels of three independent factors, theories of factorial random sampling experiments (FRSE) will be used assuming that in this three-factor sampling experiment the factor levels are random or represent a sample of possible levels.

The mathematical model used is:

$y_{ijk}$ = $\mu$ + $a_i$ + $b_j$ + $c_k$ + $ac_{ik}$ + $bc_{jk}$ + $\epsilon_{ijk}$

-   $y_{ijk}$​ represents the measured acidity level for a given batch ($i$), position ($j$), and starter culture ($k$).

-   $\mu$ represents the grand mean acidity across all conditions.

-   $a_i$ represents the effect of different batches on acidity.

-   $b_j$ represents the effect of sampling position within a batch on acidity.

-   $c_k$ represents the impact of starter culture on acidity.

-   $ac_{ik}$ represents how the impact of starter culture varies across different batches.

-   $bc_{jk}$ represents how the effect of starter culture differs depending on the position within a batch.

-   $\epsilon_{ijk}$ represents any unexplained variance not accounted for by the structured factors.

```{r}

model3 <- aov(Acidity ~ Batch + Position + Starter + Batch:Starter + Position:Starter, data = Sour_Cream)
summary(model3)


model3A <- aov(Acidity ~ Batch + Position + Starter, data = Sour_Cream)
summary(model3A)

```

The ANOVA results indicate that starter culture accounts for the largest portion of variability in acidity, followed by batch effects. The Batch:Starter interaction explains a relatively smaller portion of the variance, while position has the least impact on acidity. However, the absence of F-values and p-values in the output suggests that the model lacks sufficient degrees of freedom to properly assess statistical significance. This is likely due to the limited number of observations in the dataset, restricting the ability to formally test the significance of factor effects.

### 1.2 Estimating the variance components

The variances of the effects $a_i$, $b_j$, $c_k$, and $ac_{ik}$ are $\sigma^2_a$, $\sigma^2_b$, $\sigma^2_c$, and $\sigma^2_{ac}$, respectively. The total variability of the response can be expressed as:

$\sigma^2_y$ = $\sigma^2_a$ + $\sigma^2_b$ + $\sigma^2_c$ + $\sigma^2_{ac}$ + $\sigma^2$

-   $\sigma^2_y$ represents the total variance in the acidity across all factors and interactions.

-   $\sigma^2_a$ represents the variability in acidity explained by differences between batches.

-   $\sigma^2_b$ represents the differences in acidity based on the position within the batch.

-   $\sigma^2_c$ represents the variation in acidity due to the type of starter culture used.

-   $\sigma^2_{ac}$ represents the variability caused by how different batches respond to different starter cultures.

-   $\sigma^2$ represents all other sources of variation that are not explained by the model factors or their interactions.

```{r}

library(lme4)

model4 <- lmer(Acidity ~ 1 + (1|Batch) + (1|Position) + (1|Starter), data = Sour_Cream)
summary(model4)

# model4.1 <- lmer(Acidity ~ 1 + (1|Batch) + (1|Position) + (1|Starter) + (1|Batch:Starter), data = Sour_Cream)
# summary(model4.1)

```

The results indicate that Starter accounts for the largest portion of the variation in acidity, reflecting differences between starter cultures. This suggests that the type of starter culture used has a substantial impact on the acidity level of sour cream. Batch explains the second largest portion of variability, implying that batch-to-batch differences also influence acidity, though to a lesser extent than Starter. Position contributes the least to the variability.

The residual variation remains high. Due to the limited number of observations, $lmer()$ is unable to estimate the variation associated with the Batch:Starter interaction, which appeared to contribute only a small portion of the acidity variation based on the earlier ANOVA results. However, without statistical significance tests, it remains unclear whether these factors meaningfully contribute to explaining acidity variability.

### 1.3 Asymptotic approximate confidence intervals for variance components

```{r}

prS <- profile(fm6M <-lmer(Acidity ~ 1 + (1|Batch) + (1|Position) + (1|Starter), data = Sour_Cream, REML = FALSE))

confint(prS)

```

The confidence intervals suggest that the Batch effect (.sig01) and Starter effect (.sig03) are significantly different from zero, indicating that these factors contribute meaningfully to the variability in acidity. The Starter effect exhibits a wider range of variation compared to the Batch effect, consistent with the earlier variance component estimates.

In contrast, the Position effect (.sig02) has a lower bound of zero, suggesting that its impact on acidity may not be significantly different from zero. The residual variance (.sigma) is significantly different from zero, indicating the presence of unexplained variation.

The confidence intervals for all estimates are relatively wide, likely due to the limited number of observations in the dataset, which reduces the precision of variance component estimates.

**1.4 Model diagnostics**

```{r}

# Normality #

# Extract random effects
random_effects <- ranef(model4)

par(mfrow = c(1, 3)) # Arrange plots
qqnorm(random_effects$Batch[[1]], main = "QQ Plot - Batch Effects")
qqline(random_effects$Batch[[1]])

qqnorm(random_effects$Position[[1]], main = "QQ Plot - Position Effects")
qqline(random_effects$Position[[1]])

qqnorm(random_effects$Starter[[1]], main = "QQ Plot - Starter Effects")
qqline(random_effects$Starter[[1]])

shapiro.test(random_effects$Batch[[1]])
shapiro.test(random_effects$Position[[1]])
shapiro.test(random_effects$Starter[[1]])

# Zero mean #

summary(random_effects$Batch[[1]])
summary(random_effects$Position[[1]])
summary(random_effects$Starter[[1]])

# Constant variance #

leveneTest(Acidity ~ Batch, data = Sour_Cream)
leveneTest(Acidity ~ Position, data = Sour_Cream)
leveneTest(Acidity ~ Starter, data = Sour_Cream)
```

For the results of model4 and the estimates of variance components to be valid, the random effects Batch ($a_i$), Position ($b_j$), and Starter ($c_k$) are assumed to be normally distributed with zero means and variances $\sigma^2_a$, $\sigma^2_b$, and $\sigma^2_c$, respectively. The assessment confirms that the model meets all assumptions of normality, zero mean, and constant variance, supporting the validity of the results.

-   in first analysis, normality of the error term fails. need to fix?

-   in second analysis, significance is unclear. the variance in interaction is missing in results.

-   Slide 34

-   Is the confidence interval calculation method correct?

-   can we calculate the accuracy percentage?

-   sample size? one replicate per cell?

-   diagnostic check?

-   Tukey?

-   No need to show the dataset?
